A standardized framework for risk-based assessment of treatment effect heterogeneity in observational healthcare databases

### 2021-08-19 submission/manuscript.rmd
working on reviewer 1 comments

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012]. The
common data structure enables analyses at a very large scale. We aimed to
develop a framework for risk-based assessment of treatment effect heterogeneity
in high-dimensional observational data, which extends the existing guidelines of
the RCT setting. We implemented the framework using existing OHDSI methods for
use in the OMOP-CDM, including the patient-level prediction framework and the
population-level effect estimation framework based on new-user cohort design
[@Reps2018; @Ryan2013]. As a proof-of-concept we analyzed heterogeneity of the
effects of first-line hypertension treatment: we compared the effect of
angiotensin converting enzyme (ACE) inhibitors to beta blockers on 9 outcomes
across three different US claims databases.

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. To avoid deferentially fitting the prediction model to
patients across treatment arms, thus introducing spurious interactions with
treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.
Finally, we need to define the time horizon within which we aim to make
predictions and we also need to select the machine-learning algorithm we want to
use to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

@@

We developed separate prediction models for all efficacy outcomes (acute MI,
hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
database. More specifically, we first estimated propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices. The
prediction models were estimated on the propensity score matched (1:1) subset of
the population, using caliper of 0.2 and after excluding patients having the
outcomes any time prior to treatment initiation. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection.

>>>>

< < ADD > >

***Because of the latter issue, running analyses at scale demands a big investment
of time and effort, as researchers are forced to map their analysis plans to the
databases available to them.***

< < ADD > >

The Observational Health Data Sciences and Informatics (OHDSI) collaborative has
established an international network of data partners and researchers that aim
to bring out the value of health data through large-scale analytics by mapping
all available databases to the Observational Medical Outcomes Partnership (OMOP)
Common Data Model (CDM) [@hripcsak2015observational; @Overhage2012].

< < EDIT > >

***We aimed to develop a framework for implementing risk-based assessment of
treatment heterogeneity in high-dimensional observational data, extending the
existing guidelines of the RCT setting. Our contribution provides an
out-of-the-box solution for implementing such analyses at scale within the OHDSI
network, taking advantage of the OMOP-CDM.***

< < EDIT > >

< < OLD > >

We aimed to develop a framework for risk-based assessment of treatment effect heterogeneity
in high-dimensional observational data, which extends the existing guidelines of
the RCT setting.

< < OLD > >

We implemented the framework using existing OHDSI methods including the
patient-level prediction framework and the population-level effect estimation
framework based on new-user cohort design [@Reps2018; @Ryan2013]. As a
proof-of-concept we analyzed heterogeneity of the effects of first-line
hypertension treatment: we compared the effect of angiotensin converting enzyme
(ACE) inhibitors to beta blockers on 9 outcomes across three different US claims
databases.

@@

To generate the target cohort we pool the already defined treatment cohort and
comparator cohort. To avoid deferentially fitting the prediction model to
patients across treatment arms, thus introducing spurious interactions with
treatment [@Burke2014; @vanKlaveren2019], we develop the patient-level
prediction model in the propensity score-matched (1:1) subset of the population.

< < MOVE: ps_matching > >

< < ADD > >

***Other methods of fitting the propensity scores, such as random forest and
others (REF) can also be considered.***

< < ADD > >

Finally, we need to define the time horizon within which we aim to make
predictions and we also need to select the machine-learning algorithm we want to
use to generate patient-level predictions. Currently, the available options are
regularized logistic regression, random forest, gradient boosting machines,
decision tree, naive Bayes, K-nearest neighbors, neural network and deep
learning (convolutional neural networks, recurrent neural network and deep
nets).

@@

We developed separate prediction models for all efficacy outcomes (acute MI,
hospitalization with heart failure and hemorrhagic or ischemic stroke) in each
database.

< < MOVE: ps_matching > >

***More specifically, we first estimated propensity scores using LASSO
logistic regression and a large set of baseline covariates including
demographics, drug exposures, diagnoses, measurements and medical devices.***

< < MOVE > >

The prediction models were estimated on the propensity score matched (1:1) subset of
the population, using caliper of 0.2 and after excluding patients having the
outcomes any time prior to treatment initiation. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection. We chose a time horizon of 2
years after inclusion into the target cohort. For this demonstration, we
developed the prediction models using LASSO logistic regression with 3-fold
cross validation for hyper-parameter selection.
